---
title: "Data Harvesting"
author: "Rocío Galeote"
date: "2025-01-30"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## First part: shools list

We are going to scrape the list of Residential Schools from the NCTR archive web. We need the school name, the years it was open, the location (Town, State) and the link to their individual page. 

```{r}
rm(list=ls()) 

library(scrapex)
library(rvest)
library(httr)
library(dplyr)
library(tidyr)
library(stringr)
library(ggplot2)
library(tidyverse)

```


```{r}

# Define the URL (test with first page)
url <- "https://nctr.ca/residential-schools/"

# Read the page
page <- read_html(url)

# Extract school names
school_names <- page %>%
  html_nodes("h2.text-xl a") %>%  # Targeting <h2> with class "text-xl", then <a>
  html_text(trim = TRUE)

# Extract school links
school_links <- page %>%
  html_nodes("h2.text-xl a") %>% 
  html_attr("href")

# Extract school details (location and years)
school_details <- page %>%
  html_nodes("p.text-sm.text-gray-500") %>%  # Select <p> with class "text-sm text-gray-500"
  html_text(trim = TRUE)

# Combine into a data frame
school_data <- tibble(
  school_name = school_names,
  school_link = school_links,
  details = school_details
)

# Print the results
print(school_data)


```

```{r}

# Base URLs
base_url <- "https://nctr.ca/residential-schools/"
paged_url <- "https://nctr.ca/residential-schools/page/"

# Function to scrape a single page
scrape_school_page <- function(page_num) {
  # Handle first page differently
  url <- ifelse(page_num == 1, base_url, paste0(paged_url, page_num, "/"))
  
  # Read the page
  page <- tryCatch({
    read_html(url)
  }, error = function(e) {
    message("Error fetching page: ", url)
    return(NULL)
  })
  
  if (is.null(page)) return(tibble(school_name = NA, school_link = NA, details = NA))
  
  # Extract school names
  school_names <- page %>%
    html_nodes("h2.text-xl a") %>%  
    html_text(trim = TRUE)
  
  # Extract school links
  school_links <- page %>%
    html_nodes("h2.text-xl a") %>% 
    html_attr("href")

  # Extract school details (location and years)
  school_details <- page %>%
    html_nodes("p.text-sm.text-gray-500") %>%  
    html_text(trim = TRUE)
  
  tibble(
    school_name = school_names,
    school_link = school_links,
    details = school_details
  )
}

# Scrape all pages (1 to 15)
all_schools <- map_dfr(1:15, scrape_school_page)

# View the final data
print(all_schools)

# Save to CSV
write.csv(all_schools, "residential_schools.csv", row.names = FALSE)

```


Now we have the school name, the link to each of them, the state and years of operation. But we also want to know the religious affiliation and the table of deceased students (if they have the info), which is contained within the individual link of each school.

```{r}

library(rvest)
library(tidyverse)
library(httr)

# Define a single school URL for testing
schooltest_url <- "https://nctr.ca/residential-schools/alberta/whitefish-lake-st-andrews/"

# Read the page
page <- tryCatch({
  read_html(schooltest_url)
}, error = function(e) {
  message("Error fetching page: ", schooltest_url)
  return(NULL)
})

if (!is.null(page)) {
  # Extract Religious Entity
  religious_entity <- page %>%
    html_nodes("p.max-w-2xl span.font-normal") %>% 
    html_text(trim = TRUE)
  
  # Extract Student Memorial Table
  table_node <- page %>% html_nodes("figure.wp-block-table table")

  if (length(table_node) > 0) {
    table_data <- table_node %>% 
      html_table(fill = TRUE) %>% 
      .[[1]]  # Extract first table if multiple exist
  } else {
    table_data <- tibble(Name = NA, Date_of_Death = NA)  # Placeholder if no table
  }

  # Print results
  print(glue::glue("Religious Entity: {religious_entity}"))
  print(table_data)
}

```


Now with all

```{r}
library(rvest)
library(tidyverse)
library(httr)

# Function to scrape a single school's details using its link

scrape_school_details <- function(school_link) {
  # Ensure the full URL is formed correctly
  full_url <- ifelse(grepl("^https?://", school_link), school_link, paste0("https://nctr.ca", school_link))
  
  # Read the page
  page <- tryCatch({
    read_html(full_url)
  }, error = function(e) {
    message("Error fetching page: ", full_url)
    return(NULL)
  })
  
  if (is.null(page)) return(tibble(school_link = school_link, religious_entity = NA, student_data = NA))
  
  # Extract Religious Entity
  religious_entity <- page %>%
    html_nodes("p.max-w-2xl span.font-normal") %>%
    html_text(trim = TRUE)
  
  # Extract Student Memorial Table
  table_node <- page %>% html_nodes("figure.wp-block-table table")
  
  if (length(table_node) > 0) {
    table_data <- table_node %>% 
      html_table(fill = TRUE) %>% 
      .[[1]]  # Extract first table if multiple exist
  } else {
    table_data <- tibble(Name = NA, Date_of_Death = NA)  # Placeholder if no table
  }
  
  tibble(
    school_link = school_link,
    religious_entity = ifelse(length(religious_entity) > 0, religious_entity, NA),
    student_data = list(table_data)  # Store tables in a list-column
  )
}

# Apply the scraping function to all school links from `all_schools`
detailed_school_data <- map_dfr(all_schools$school_link, scrape_school_details)

# View the results
print(detailed_school_data)

```

We join them by school_link

```{r}
# Ensure dplyr is loaded
library(dplyr)

# Join the two dataframes by the `school_link` column
combined_data <- left_join(all_schools, detailed_school_data, by = "school_link")

# View the combined data
print(combined_data)

```

We want to separate the details of each school into: school type, location and years_operation, and then remove the original column

```{r}
library(dplyr)
library(stringr)

combined_data <- combined_data %>%
  mutate(
    # Extract School Type (before " - ")
    school_type = str_extract(details, "^[^ -]+"),

    # Extract Location (between " - " and the year in parentheses)
    location = str_extract(details, "(?<= - ).*(?= \\()"),

    # Extract Years of Operation (in parentheses)
    years_operation = str_extract(details, "\\(.*\\)")
  ) %>%
  # Remove the original 'details' column
  select(-details)

# View the updated combined data with the new columns
print(combined_data)
```


We need to clean the years_operation column, there is duplicate data. Also, there are weird spaces within the numbers and we need to standardize them. We chose to remove the parentheses too:

```{r}
library(dplyr)
library(stringr)

combined_data <- combined_data %>%
  mutate(
    # Extract only the last set of parentheses with years
    years_operation = str_extract(years_operation, "\\([^()]*\\)$") %>%
      str_replace_all("[()]", "") %>%  # Remove parentheses
      str_trim() %>%  # Remove any leading or trailing spaces
      str_replace_all("\\s*[–—-]\\s*", "-")  # Standardize dashes to "-"
  )

# View the cleaned data
print(combined_data)
```


We're interested in seeing how many years each school was operating, so we extract the years from the column year_operation, we set them as numeric and we compute the substraction:

```{r}

library(dplyr)
library(stringr)

combined_data <- combined_data %>%
  mutate(
    # Extract the first and last year as numeric values
    start_year = as.numeric(str_extract(years_operation, "^\\d{4}")),
    end_year = as.numeric(str_extract(years_operation, "\\d{4}$")),
    
    # Calculate years in operation (if both values exist)
    years_active = ifelse(!is.na(start_year) & !is.na(end_year),
                          end_year - start_year,
                          NA)
  ) %>%


# View the cleaned data
print(combined_data)

```



## Second part: Interactive map

Next, we go to the Arcgis map, because Selenium can't scrape an iframe that leads to another website. We detect how many markers there are in layer 9, we remove any spinner hidden in the code, locate the popup when clicking on a marker and returning the info on Title, Longitude and Latitude.


```{r}
library(RSelenium)
library(stringr)

# Connect to Selenium running in Docker
remDr <- remoteDriver(
  remoteServerAddr = "localhost",
  port = 4449,
  browserName = "firefox"
)

remDr$open()

# Navigate to the ArcGIS map
remDr$navigate("https://www.arcgis.com/apps/webappviewer/index.html?id=0cc12b7f9f434dbaa7c2815aea84606e/")
Sys.sleep(5)  # Wait for page to load

# Find all markers in layer 9 (escape ID properly)
layer_9_markers <- remDr$findElements(using = "css selector", "#\\31 94bd5e080b-layer-9_layer > image")
print(paste("Layer 9 Markers found:", length(layer_9_markers)))


# 🚀 **Remove Hidden Spinner (Preemptive)**
remDr$executeScript("
  Array.from(document.querySelectorAll('.spinner.hidden')).forEach(el => el.remove());
")
Sys.sleep(5) 

for (marker in layer_9_markers) {
  Sys.sleep(2)  # Wait a bit before clicking
  marker$clickElement()
  
  # Ensure no popups are blocking
  close_buttons <- remDr$findElements(using = "css selector", "div.esriMobileNavigationBar img")
  if (length(close_buttons) > 0) {
    close_buttons[[1]]$clickElement()
    Sys.sleep(2)
  }

  # Remove spinner if it's back
  spinners <- remDr$findElements(using = "css selector", ".spinner.hidden")
  if (length(spinners) > 0) {
    remDr$executeScript("document.querySelectorAll('.spinner.hidden').forEach(el => el.remove());")
    Sys.sleep(1)
  }

  # Scroll to marker
  remDr$executeScript("arguments[0].scrollIntoView();", list(marker))
  Sys.sleep(1)

  # 🚀 **Remove Hidden Spinner Again**
  remDr$executeScript("
    Array.from(document.querySelectorAll('.spinner.hidden')).forEach(el => el.remove());
  ")

  # Locate the popup (single instance, since it updates content)
  popup <- remDr$findElement(using = "css selector", "div.esriPopupWrapper")

  remDr$screenshot(display = TRUE)  # Show image in RStudio Viewer

  # Extract the title from the header
  title_element <- popup$findChildElement(using = "css selector", ".header")
  title <- title_element$getElementText()[[1]]

  # Find all table rows in the popup
  rows <- popup$findChildElements(using = "css selector", ".attrTable tr")

  latitude <- "N/A"
  longitude <- "N/A"

  for (row in rows) {
    # Get columns (td elements) in each row
    cols <- row$findChildElements(using = "css selector", "td")

    if (length(cols) >= 2) {
      name_element <- cols[[1]]  # First column (attribute name)
      value_element <- cols[[2]] # Second column (attribute value)

      name_text <- name_element$getElementText()[[1]]
      value_text <- value_element$getElementText()[[1]]

      if (grepl("Latitude", name_text, ignore.case = TRUE)) {
        latitude <- value_text
      } else if (grepl("Longitude", name_text, ignore.case = TRUE)) {
        longitude <- value_text
      }
    }
  }

  print(paste("Title:", title, "Latitude:", latitude, "Longitude:", longitude))

# ✅ **Close the Popup Before Moving to Next Marker**
tryCatch({
  close_button <- remDr$findElement(using = "css selector", "div.titleButton.close")
  close_button$clickElement()
  Sys.sleep(2)  # Allow time for the popup to close
}, error = function(e) {
  print("Popup close button not found or not clickable, moving on...")
})

}

  # Save data
school_data <- data.frame(Name = character(), Latitude = numeric(), Longitude = numeric(), stringsAsFactors = FALSE)

# Close the session
remDr$close()


```


This next part i dont know what its for, tbh

```{r}
library(RSelenium)
library(stringr)

# Connect to Selenium running in Docker
remDr <- remoteDriver(
  remoteServerAddr = "localhost",
  port = 4449,
  browserName = "firefox"
)

remDr$open()

# Navigate to the ArcGIS map
remDr$navigate("https://www.arcgis.com/apps/webappviewer/index.html?id=0cc12b7f9f434dbaa7c2815aea84606e/")
Sys.sleep(5)  # Wait for page to load

# Find all markers in layer 9 (escape ID properly)
layer_9_markers <- remDr$findElements(using = "css selector", "#\\31 94bd5e080b-layer-9_layer > image")
print(paste("Layer 9 Markers found:", length(layer_9_markers)))

# ✅ Initialize dataframe BEFORE the loop
school_data <- data.frame(Name = character(), Latitude = numeric(), Longitude = numeric(), stringsAsFactors = FALSE)

# 🚀 **Remove Hidden Spinner (Preemptive)**
remDr$executeScript("
  Array.from(document.querySelectorAll('.spinner.hidden')).forEach(el => el.remove());
")
Sys.sleep(5) 

for (marker in layer_9_markers) {
  Sys.sleep(2)  # Wait a bit before clicking
  marker$clickElement()
  
  # Ensure no popups are blocking
  close_buttons <- remDr$findElements(using = "css selector", "div.esriMobileNavigationBar img")
  if (length(close_buttons) > 0) {
    close_buttons[[1]]$clickElement()
    Sys.sleep(2)
  }

  # Remove spinner if it's back
  spinners <- remDr$findElements(using = "css selector", ".spinner.hidden")
  if (length(spinners) > 0) {
    remDr$executeScript("document.querySelectorAll('.spinner.hidden').forEach(el => el.remove());")
    Sys.sleep(1)
  }

  # Scroll to marker
  remDr$executeScript("arguments[0].scrollIntoView();", list(marker))
  Sys.sleep(1)

  # 🚀 **Remove Hidden Spinner Again**
  remDr$executeScript("
    Array.from(document.querySelectorAll('.spinner.hidden')).forEach(el => el.remove());
  ")

  # Locate the popup (single instance, since it updates content)
  popup <- remDr$findElement(using = "css selector", "div.esriPopupWrapper")


  # Extract the title from the header
  title_element <- popup$findChildElement(using = "css selector", ".header")
  title <- title_element$getElementText()[[1]]

  # Find all table rows in the popup
  rows <- popup$findChildElements(using = "css selector", ".attrTable tr")

  latitude <- "N/A"
  longitude <- "N/A"

  for (row in rows) {
    # Get columns (td elements) in each row
    cols <- row$findChildElements(using = "css selector", "td")

    if (length(cols) >= 2) {
      name_element <- cols[[1]]  # First column (attribute name)
      value_element <- cols[[2]] # Second column (attribute value)

      name_text <- name_element$getElementText()[[1]]
      value_text <- value_element$getElementText()[[1]]

      if (grepl("Latitude", name_text, ignore.case = TRUE)) {
        latitude <- value_text
      } else if (grepl("Longitude", name_text, ignore.case = TRUE)) {
        longitude <- value_text
      }
    }
  }

  # ✅ **Append the extracted data to the dataframe**
  school_data <- rbind(school_data, data.frame(Name = title, Latitude = latitude, Longitude = longitude, stringsAsFactors = FALSE))

  # ✅ **Close the Popup Before Moving to Next Marker**
  tryCatch({
    close_button <- remDr$findElement(using = "css selector", "div.titleButton.close")
    close_button$clickElement()
    Sys.sleep(2)  # Allow time for the popup to close
  }, error = function(e) {
    print("Popup close button not found or not clickable, moving on...")
  })
}

# ✅ Print final dataframe
print(school_data)

# ✅ Save dataframe to CSV file
write.csv(school_data, "school_data.csv", row.names = FALSE)

# Close the session
remDr$close()

```


Works but we have 52 missing data when we shouldn't. We try a new approach scrolling the map. Printing each result in the console and screenshoting every time it scrolls.

```{r}
library(RSelenium)
library(stringr)

# Connect to Selenium running in Docker
remDr <- remoteDriver(
  remoteServerAddr = "localhost",
  port = 4449,
  browserName = "firefox"
)

remDr$open()

# Navigate to the ArcGIS map
remDr$navigate("https://www.arcgis.com/apps/webappviewer/index.html?id=0cc12b7f9f434dbaa7c2815aea84606e/")
Sys.sleep(5)  # Wait for page to load

# Remove hidden spinners
remDr$executeScript("
  Array.from(document.querySelectorAll('.spinner.hidden')).forEach(el => el.remove());
")
Sys.sleep(5) 

# Function to take a screenshot
take_screenshot <- function(filename) {
  screenshot_file <- paste0(filename, ".png")
  remDr$screenshot(file = screenshot_file)
  print(paste("📸 Screenshot saved:", screenshot_file))
}

# Function to move (pan) the map
move_map <- function(x_offset, y_offset, step_name) {
  print(paste("🗺️ Moving map:", step_name))
  remDr$executeScript("
    let map = document.querySelector('.esriViewRoot'); 
    map.scrollBy(arguments[0], arguments[1]);", 
    list(x_offset, y_offset)
  )
  Sys.sleep(3)  # Allow time for map to load
  take_screenshot(paste0("map_moved_", step_name))
}

# Define scroll movements
movements <- list(
  list(x = 500, y = 0, name = "Right"),
  list(x = -1000, y = 0, name = "Left"),
  list(x = 500, y = 500, name = "Down"),
  list(x = 500, y = -1000, name = "Up")
)

# Function to scrape markers
scrape_markers <- function() {
  layer_9_markers <- remDr$findElements(using = "css selector", "#\\31 94bd5e080b-layer-9_layer > image")
  print(paste("🔍 Total markers found:", length(layer_9_markers)))

  for (marker in layer_9_markers) {
    Sys.sleep(2)
    marker$clickElement()
    
    # Remove spinner if it reappears
    spinners <- remDr$findElements(using = "css selector", ".spinner.hidden")
    if (length(spinners) > 0) {
      remDr$executeScript("document.querySelectorAll('.spinner.hidden').forEach(el => el.remove());")
      Sys.sleep(1)
    }

    Sys.sleep(2)
    popups <- remDr$findElements(using = "css selector", "div.esriPopupWrapper")

    if (length(popups) == 0) {
      print("⚠️ Popup not found. Skipping marker.")
      next
    }

    popup <- popups[[1]]
    title_element <- popup$findChildElement(using = "css selector", ".header")
    title <- title_element$getElementText()[[1]]

    # Find all table rows in the popup
    rows <- popup$findChildElements(using = "css selector", ".attrTable tr")

    latitude <- "N/A"
    longitude <- "N/A"

    for (row in rows) {
      cols <- row$findChildElements(using = "css selector", "td")
      if (length(cols) >= 2) {
        name_text <- cols[[1]]$getElementText()[[1]]
        value_text <- cols[[2]]$getElementText()[[1]]

        if (grepl("Latitude", name_text, ignore.case = TRUE)) {
          latitude <- value_text
        } else if (grepl("Longitude", name_text, ignore.case = TRUE)) {
          longitude <- value_text
        }
      }
    }

    print(paste("✅ Title:", title, "Latitude:", latitude, "Longitude:", longitude))

    # Close the popup
    tryCatch({
      close_button <- remDr$findElement(using = "css selector", "div.titleButton.close")
      close_button$clickElement()
      Sys.sleep(2)
    }, error = function(e) {
      print("Popup close button not found. Skipping closure.")
    })
  }
}

# 🚀 **Scrape first pass**
scrape_markers()

# 🚀 **Scroll the map in all directions**
for (move in movements) {
  move_map(move$x, move$y, move$name)
  scrape_markers()  # Try scraping again after each movement
}

# 🚀 **Final pass to ensure all markers were scraped**
scrape_markers()

# Save data
write.csv(scraped_markers, "scraped_markers.csv", row.names = FALSE)

# Close the session
remDr$close()


```


```{r}
library(RSelenium)
library(stringr)

# Connect to Selenium running in Docker
remDr <- remoteDriver(
  remoteServerAddr = "localhost",
  port = 4449,
  browserName = "firefox"
)

remDr$open()

# Navigate to the ArcGIS map
remDr$navigate("https://www.arcgis.com/apps/webappviewer/index.html?id=0cc12b7f9f434dbaa7c2815aea84606e/")
Sys.sleep(5)  # Wait for page to load


# Locate the Map Element
mapElement <- remDr$findElement(using = "xpath", value = "//*[@id='map']")  # Adjust the XPath as necessary

# Define Drag Offsets for each direction
drag_offsets <- list(
    list(startX = 0, startY = 0, endX = 500, endY = 0),   # Right
    list(startX = 500, startY = 0, endX = 500, endY = 500), # Down
    list(startX = 500, startY = 500, endX = 0, endY = 500), # Left
    list(startX = 0, startY = 500, endX = 0, endY = 0)      # Up
)

# Loop Through Offsets and Take Screenshots
for (i in seq_along(drag_offsets)) {
    offset <- drag_offsets[[i]]
    
    # Take Initial Screenshot
    remDr$screenshot(file = paste0("before_drag_", i, ".png"))
    print(paste("📸 Screenshot saved: before_drag_", i, ".png", sep = ""))

    # Define the JavaScript for dragging
    drag_script <- "
    var map = arguments[0];
    var mouseDownEvent = new MouseEvent('mousedown', {
        view: window,
        bubbles: true,
        cancelable: true,
        clientX: arguments[1],
        clientY: arguments[2]
    });
    map.dispatchEvent(mouseDownEvent);

    var mouseMoveEvent = new MouseEvent('mousemove', {
        view: window,
        bubbles: true,
        cancelable: true,
        clientX: arguments[3],
        clientY: arguments[4]
    });
    map.dispatchEvent(mouseMoveEvent);

    var mouseUpEvent = new MouseEvent('mouseup', {
        view: window,
        bubbles: true,
        cancelable: true,
        clientX: arguments[3],
        clientY: arguments[4]
    });
    map.dispatchEvent(mouseUpEvent);
    "

    # Execute the drag script with offsets
    remDr$executeScript(drag_script, list(mapElement, offset$startX, offset$startY, offset$endX, offset$endY))

    # Take Screenshot After Dragging
    remDr$screenshot(file = paste0("after_drag_", i, ".png"))
    print(paste("📸 Screenshot saved: after_drag_", i, ".png", sep = ""))
}


remDr$close()



```

